Concurrent Programming with CXL

### 2020-06-07 book/paper.tex
Rough draft of split binary semaphores

We have seen solutions to the reader/writer problem and the producer/consumer
problem that do no busy-wait, but they are quite different.  It would be nice
to have a general technique for all CCR problems.
The ``Split Binary Semaphore'' is such a technique, originally proposed by
Tony Hoare and popularized by Dijkstra.

>>>>

We have seen solutions to the reader/writer problem and the producer/consumer
problem that do no busy-wait, but they are quite different.  It would be nice
to have a general technique for all CCR problems.
The ``Split Binary Semaphore'' (SBS) is such a technique, originally proposed by
Tony Hoare and popularized by Edsger Dijkstra.
SBS is an example of what one could call a ``baton-passing'' technique:
when a process has the
baton and is about to leave the critical region, giving up the baton, it first
checks to see if there are any processes that are waiting and can now run.
If so, it passes the baton directly to one of those processes.

We will illustrate the technique using the reader/writer problem.
Figure~\ref{fig:RWsplitsema} shows the
\texttt{rlock\_acquire}, \texttt{rlock\_release},
\texttt{wlock\_acquire}, and \texttt{wlock\_release} methods implemented using
SBS.

Step 1 is to enumerate all waiting conditions.  In the case of the reader/writer
problem, there are two: a process that wants to read may have to wait for a
writer to leave the critical regiion, while a process that wants to write may
have to wait until all readers have left the critical region.  If there are $N$
waiting conditions, the SBS technique will use $N+1$ binary semaphores: one for
each waiting condition and an extra one that processes use when they first try
to enter the critical region and do not yet know if they need to wait.
We shall call this the \texttt{mutex}.

An important invariant of the SBS technique is that the sum of the $N+1$ semaphores
must always be 1 if there is no process in the CCR and 0 if there is
a process in the CCR.  As it were, these semaphores taken together are emulating a
single binary semaphore.
Initially, \texttt{mutex} is 1 and the other $N$ semaphores
are all 0.  To maintain the invariant, each process alternates \texttt{P} and
\texttt{V} operations, starting with \texttt{P~\&(mutex)} when it tries to
enter the CCR, and ending with a \texttt{V} operation on one of the semaphores
(yet to be determined).

Another important part of the SBS technique is to keep careful track of the
state.  In the case of the reader/writer lock, the state consists:
\begin{itemize}
\item \texttt{nreadersEntered}: the number of readers in the critical region;
\item \texttt{nwritersEntered}: the number of writers in the critical region;
\item \texttt{nreadersWaiting}: the number of readers waiting to enter;
\item \texttt{nwritersWaiting}: the number of writers waiting to enter.
\end{itemize}
Each of the
\texttt{rlock\_acquire}, \texttt{rlock\_release},
\texttt{wlock\_acquire}, and \texttt{wlock\_release} methods must maintain
this state.
To wit, check out \texttt{rlock\_acquire},  It first checks to see if there
are any writers.  If so, it increments \texttt{nreadersWaiting}.  If not,
it increments \texttt{nreadersEntered}.
Either way, it calls the method \texttt{V\_one()}.

Note that the four
reader/writer methods never call \texttt{V} on a specific binary semaphore:
\texttt{V\_one()} is the baton passing function that selects which of the
binary semaphores to increment.
Method \texttt{V\_one()} first checks to see if there are any readers or
writers waiting.  If there are readers waiting and there are no writers
in the critical section, it increments the semaphore associated with
readers waiting.  If there are writers waiing and there are no readers
nor writers in the critical section, it increments the semaphore associated
with writers waiting.  If both conditions hold, it could increment either
one---it would not matter.  However, it can only increment one of the semaphores.
If none of the conditions hold, the \texttt{V\_one} increments \texttt{mutex}.

Let us return now to \texttt{rlock\_acquire}, in the case that the process
found there is a writer in the critical region.  After \texttt{V\_one()} returns,
it invokes calls \texttt{P~\&(rsema)}, blocking on the semaphore associated
with waiting readers.  When later some other process passes the baton to
it by incrementing \texttt{rsema}), it updates the state by decrementing
\texttt{nreadersWaiting} and incrementing \texttt{nreadersEntered}.  It
then invokes \texttt{V\_one()} one more time.

As an example, consider the case where there is a writer in the critical
section and there are two readers waiting.  Let us see what happens when
the writer calls \texttt{wlock\_release}:
\begin{enumerate}
\item After procuring \texttt{mutex},
the sum of the three semaphores is 0.  The writer then decrements
\texttt{nwritersEntered} and calls \texttt{V\_one()}.
\item \texttt{V\_one()} finds that there are no writers in the critical region
and there are two readers waiting.  It therefore vacates \texttt{rsema}.
The sum of the semaphores is now 1, because \texttt{rsema} is 1.  This
guarantees that only a waiting reader, procuring \texttt{rsema}, can enter
the critical region.
\item When it does, the reader decrements \texttt{nreadersWaiting}
from 2 to 1, and increments \texttt{nReadersEntered} from 0 to 1.
The reader finally calls \texttt{V\_one()}.
\item Again, \texttt{V\_one()} finds that there are no writers and
that there are readers waiting, so again it vacates \texttt{rsema},
passing the baton to the one remaining waiting reader.
\item The reader decrements \texttt{nreadersWaiting} from 1 to 0 and
increments \texttt{nReadersEntered} from 1 to 2.
\item Finally, the reader  calls \texttt{V\_one()}.
\texttt{V\_one()} does not find any processes waiting,
and so it vacates \texttt{mutex}.
\end{enumerate}
